\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-19 introduction.tex
pass on lang-design sec

Integrating family polymorphism into a dependent type theory with
inductive types poses significant technical challenges, however.
As we analyze, inductive types, definitional equalities, and logical
consistency, are all inimical to the kind of family polymorphism
found in existing language designs.
Thus, our core contributions include design recipes for dealing with
these issues and strong metatheoretical guarantees on consistency and
canonicity.

\paragraph{Contributions} To our knowledge, we offer the first
language design that enables extensible metatheory mechanization
in a higher-order, dependent type theory with inductive types.

@@

\item We capture the new language mechanism formally by extending
Martin–Löf type theory with family polymorphism and extensible inductive
types (\cref{sec:metatheory}).
We derive strong metatheoretical results including on consistency and
canonicity.
We also formalize the translation implemented by the Coq plugin.

\item We present case studies of using our Coq plugin to mechanize
language metatheories.
They show how our linguistic approach naturally solves the EP and
enables a high degree of reusability and extensibility
for proof engineering.

>>>>

Integrating family polymorphism into a dependent type theory with
inductive types poses significant technical challenges, however.
As we analyze, pillars of such a dependent type theory—including
inductive types, definitional equalities, and logical consistency—are
all inimical to the kind of extensibility and family polymorphism
found in existing OO language designs.
Thus, our contributions include novel design recipes for dealing with
these challenges and strong meta\-theoretical guarantees on the
underlying logic.

\paragraph{Contributions} To our knowledge, we offer the first
language-based solution that enables extensible metatheory mechanization
in a higher-order, dependent type theory with inductive types.

@@

\item We capture the new language mechanism formally by extending
Martin–Löf type theory with family polymorphism and extensible inductive
types (\cref{sec:metatheory}).
We derive strong meta\-theoretical results including on consistency and
canonicity.
We also formalize the translation implemented by the Coq plugin.

\item We present case studies of using our Coq plugin to mechanize
language metatheories.
They show how our language design naturally solves the EP and
enables a high degree of reuse and extensibility
for proof engineering.

### 2022-07-19 lang-design.tex
Alter the simple negation example into the complicated STLC example In Coqimpl, implementation of extensible inductive type. Please check
pass on lang-design sec

\begin{figure}[!htb]
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
 FInductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
 FInductive term : Set := 
  | var : id -> term 
  | lam : id -> term -> term ...
 Family subst_internal.
  Final Field var := ..
  Final Field lam := ..
   ...
 EndFamily.
 FRecursor subst
  about tm motive ..
  using subst_internal.
 ...
 FInductive has_type 
  : context -> term -> type := ...
 FTheorem subst_lemma 
  about has_type 
  motive ...
 Proof.  ...
 Qed.
 FInductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC_bool extends STLC.
 Extend FInductive ty : Set :=
  | bool : ty.

@@

% Inheritance here
\textbf{Inheritance on Fields.} "Family A extends B" has to appear at the top level (not inside any Family) and will start to construct the children "Family A" given the parent "Family B". Every field inside the inheriting family will be one of the three kinds of ``inheritance''---extending, overriding, and inheriting. The syntax of extending a new field is to directly use "Field", while overriding requires a prefix of "Override" before "Field f : .." and "Inherit f" will inherits field "f" from the parent.

To resolve \ref{chg:consistency}, we make sure when inheriting or overriding on "f" happens, the context of the field "f" is also ``inherited''. Thus the mentioned informal example in \ref{chg:consistency} will not apply here because the overriding "f" is also in the context where "self" is empty thus "self.g" doesn't type-check. Thus, before we inherit "g", the dependee "f" needs to be inherited.

@@

% Talk about Final
\textbf{Final as non-overridable.} Every field is by default overridable, reminding us \ref{chg:definition-relevant-reasoning}.  To resolve it, we provide the keyword "Final" to allow the user to specify \textit{non-overridability} and give the user the opportunity to choose from overriding and exposing the definition.

Take the field "c" of "Family I" in \cref{fig:plugin-example3} as the
example. We can see that the first attempt of proving "a = 1" fails.
This should fail---because we want to retain the possibility
to override "a" and inheriting "c" in the future, but they are conflicting, since "c" witnesses "a" fix to "1". On the other hand, once we decorate it upon "a'", the second attempt \mintinline{Coq}{c : a' = 1 := eq_refl} will succeed. Contrary to "a", our "a'" is not overridable in "Family I2", mediating the \ref{chg:definition-relevant-reasoning}.

This example might sound trivial and contrived. Let's look at the
example field binary operation "bop" initially set to be addition.\YZ{This 'add' field is no longer there.}\EDJreply{Thanks. Fixed} Now that we only expose the type of "bop", we
cannot prove its commutativity in "comm_bop" (because we don't have
access to its concrete definition) but we can override "bop" with
subtraction function in "Family I2". This makes sense because we
shouldn't (have the ability to) inherit commutativity property for subtraction
anyway. This is contrary to how "a'" is declared "Final", the second "c"
can prove that it's equal to 1 but "a'" is not overridable in the
"Family I2".

@@

To resolve \ref{chg:extensible-inductive-type}, our first step is to \textit{decouple} the definition of recursor handlers and the checking of exhaustiveness. The users will define recursive handlers for one inductive type by aggregating the (function) handlers into one family. Then
"FRecursor" will be used to construct a recursor by refering to the handler family.
\mintinline{Coq}{FRecursor recur about T motive f } "using"
"handlers" will generate a new field "recur" that is really doing computation
(using elimination principle of the inductive type "T") with motive "f". The handlers
are located in the family "handlers", similar to a large pattern-matching statement. This way, the reuse and the extension of the case handlers will be delegated by the inheritance of the family. Once recursor is created it will look like any other function fields(e.g. "subst" in \cref{fig:STLC-example2}).

"FTheorem" has a similar functionality but is expected to be used for theorem proving---the handler and the family will be automatically constructed during proving.

>>>>

\begin{figure}[!htb]
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
 FInductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
 FInductive term : Set := 
  | tm_var : id -> term 
  | tm_abs : id -> term -> term ...
 Family subst_internal.
  Final Field tm_var := ..
  Final Field tm_abs := ..
   ...
 EndFamily.
 FRecursor subst
  about tm motive ..
  using subst_internal.
 ...
 FInductive has_type 
  : context -> term -> type := ...
 FTheorem subst_lemma 
  about has_type 
  motive ...
 Proof.  ...
 Qed.
 FInductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC_bool extends STLC.
 Extend FInductive ty : Set :=
  | bool : ty.

@@

% Inheritance here
\textbf{Inheritance on Fields.} "Family A extends B" has to appear at
the top level (not inside any Family) and will start to construct the
children "Family B" given the parent "Family A". Every field inside the
derived family will be one of the three kinds of
``inheritance''---extending, overriding, and inheriting.
The syntax of extending a new field is to directly use "Field", while
overriding requires a prefix of "Override" before "Field f : .." and
"Inherit f" will inherit field "f" from the parent.

To resolve \ref{chg:consistency}, we make sure when inheriting or
overriding on "f" happens, the context of the field "f" is also
``inherited''. Thus the mentioned informal example in
\ref{chg:consistency} will not apply here, because the overriding "f" is
in a context where "self" is empty thus "self.g" doesn't
type-check. Thus, before we inherit "g", the dependee "f" needs to be
inherited.

@@

% Talk about Final
\textbf{Final as non-overridable.} Every field is by default
overridable, reminding us of \ref{chg:definition-relevant-reasoning}.  To
resolve it, we provide the keyword "Final" to allow the user to specify
\textit{non-overridability} and give the user the opportunity to choose
from overriding and exposing the definition.

Take the field "c" of "Family I" in \cref{fig:plugin-example3} as an 
example. We can see that the first attempt of proving "a = 1" fails.
This should fail---because we want to retain the possibility
to override "a" and inheriting "c" in the future, but they are
conflicting, since "c" witnesses that "a" is fixed to "1". On the other
hand, once we decorate \lstinline{a'} as final, the second attempt
\lstinline{c : a' = 1 := eq_refl} will succeed. Contrary to "a",
our \lstinline{a'} is not overridable in "Family I2", mediating the
\ref{chg:definition-relevant-reasoning}.

This example might sound trivial and contrived. Let's look at the
example field binary operation "bop" initially set to be addition.
Now that we only expose the type of "bop", we
cannot prove its commutativity in "comm_bop" (because we don't have
access to its concrete definition) but we can override "bop" with
subtraction function in "Family I2". This makes sense because we
shouldn't (have the ability to) inherit commutativity property for
subtraction anyway. This is contrary to how \lstinline{a'} is declared
"Final": the second "c" can prove that it's equal to 1 but
\lstinline{a'} is not overridable in the "Family I2".

@@

To resolve \ref{chg:extensible-inductive-type}, our first step is to
\textit{decouple} the definition of recursor handlers and the checking
of exhaustiveness. The users will define recursive handlers for one
inductive type by aggregating the (function) handlers into one family.
Then
"FRecursor" will be used to construct a recursor by referring to the handler family.
\mintinline{Coq}{FRecursor recur about T motive f } "using"
"handlers" will generate a new field "recur" that is really doing computation
(using elimination principle of the inductive type "T") with motive "f". The handlers
are located in the family "handlers", similar to a large pattern-matching statement. This way, the reuse and the extension of the case handlers will be delegated by the inheritance of the family. Once recursor is created it will look like any other function fields(e.g. "subst" in \cref{fig:STLC-example2}).

"FTheorem" has a similar functionality but is expected to be used for
theorem proving---the handler and the family will be automatically
constructed during proving.\YZ{Cool. Can the same be done to FRecursor?}
